---
title: "Main steps in extracting win-loss measures"
subtitle: "Step-by-step vs. all-in-one"
execute:
  eval: false
bibliography: references.bib
---

This document outlines two approaches to calculating win-loss statistics
using KM and summary data:

(1) A **step-by-step** approach calling the following functions in turn:

- `prepare_km_data()` to read and clean digitized KM data.
- `merge_endpoints()` to align OS and PFS on a common time grid.
- `compute_increments()` to calculate small increments/decrements for OS/PFS needed
to calculate win-loss probabilities [@oakes:2016].
- `compute_followup()` to derive total follow-up times from at-risk tables.
- `compute_theta()` to compute association parameters ($\theta$) between the two endpoints.
- `compute_win_loss()` to calculate final win/loss probabilities with or without $\theta$.

(2) An **all-in-one** approach using `run_win_loss_workflow()` as a single wrapper that runs all of the above steps at once.

---

## Step-by-step approach

### 1. Data preparation: `prepare_km_data()`

**Purpose:**  Read and clean digitized KM data.

**Main Inputs:**

- `km_data`: A data frame in wide format, typically read from a CSV containing digitized times and survival probabilities for multiple groups.
- `time_cols`: Character vector of column names that contain the time values for each group (e.g. `"time_placebo", "time_camr"`).
- `surv_cols`: Character vector of column names that contain the survival probabilities for each group (e.g. `"surv_placebo", "surv_camr"`).
- `ref`: (Optional) A character string indicating which group is the reference (control).
- `group_labels`: (Optional) A character vector to rename the group factors (must match the number of unique groups).  
- `zero_time`: (Logical) If `TRUE`, adds (time=0, surv=1) for each group.

**Output:**  A tibble with columns:

- `group`: Factor indicating treatment/control groups.
- `time`: Time points (including 0 if `zero_time=TRUE`).
- `surv`: Survival probability at each time (forced non-increasing by group).
- Optionally reordered factor levels if `ref` is used.

**Example Usage:**

```{r}
# For OS data
os_long <- prepare_km_data(
  km_data      = os_data,
  time_cols    = c("time_placebo", "time_camr"),
  surv_cols    = c("surv_placebo", "surv_camr"),
  ref          = "placebo",
  group_labels = c("Placebo", "Camrelizumab"),
  zero_time    = TRUE
)
# For PFS data
pfs_long <- prepare_km_data(
  km_data      = pfs_data,
  time_cols    = c("time_placebo", "time_camr"),
  surv_cols    = c("surv_placebo", "surv_camr"),
  ref          = "placebo",
  group_labels = c("Placebo", "Camrelizumab"),
  zero_time    = TRUE
)
```

### 2. Merge endpoints: `merge_endpoints()`

**Purpose:** Combine two sets of KM data (e.g. OS and PFS) onto a common time grid.

**Main Inputs:**

-   `os_data_long`: A long-format tibble, typically from prepare_km_data(), containing OS.
-   `pfs_data_long`: A long-format tibble, similarly structured, containing PFS.
-   `extra_times`: (Optional) Numeric vector of additional time points to include in the merged grid.

**Output:** A tibble with columns:

-   `group`: Group identifier (same factor levels as inputs).
-   `time`: Combined and sorted time points from OS and PFS.
-   `os, pfs`: Interpolated OS/PFS probabilities at each time.

**Example Usage:**

```{r}
df_merged <- merge_endpoints(os_long, pfs_long)
```

### 3. Compute increments: `compute_increments()`

**Purpose:** Compute small decrements in survival probabilities for OS and PFS
in preparation for calculation of win-loss estimands [@oakes2016].

**Main Inputs:**

-   `df_km`: A long-format tibble with columns `group`, `time`, `os` $(\hat S(t))$, `pfs` $(\hat S^*(t))$ (e.g. from `merge_endpoints()`).

**Output:** A tibble with additional columns:

-   `os_p`: Previous (lagged) OS $(\hat S(t-))$.
-   `dos`: Decrement in OS (`os_p` - `os`) $(\hat S(t-) - \hat S(t))$.
-   `pfs_p`: Previous (lagged) PFS $(\hat S^*(t-))$.
-   `dpfs`: Decrement in PFS (`pfs_p` - `pfs`) $(\hat S^*(t-) - \hat S^*(t))$.
-   `dLpfs`: Hazard increment for PFS $\mathrm{d}\hat\Lambda^*(t)$.

**Example Usage:**

```{r}
df_increments <- compute_increments(df_merged)
```

### 4. Compute follow-up times: `compute_followup()`

**Purpose:** Compute the total person-time (follow-up) by endpoint and group from a risk table.

**Main Inputs:**

-   `risk_table`: Data frame of at-risk numbers in wide format, with columns:
    -   A `time` column.
    -   An `endpoint` column (e.g., "os"/"pfs").
    -   Two columns for group-specific at-risk counts.
-   `time_col`: (Default = "time") Name of the time column.
-   `endpoint_col`: (Default = "endpoint") Name of the endpoint column.
-   `group_cols`: Character vector specifying which columns hold the at-risk counts.
-   `group_labels`: (Optional) Renames the factor levels after pivoting wide $\to$ long.

**Output:** A tibble with columns:

-   `endpoint`: E.g. "os" or "pfs".
-   `group`: Group label (factor).
-   `total_followup`: Person-time computed for that group/endpoint.


### 5. Compute association parameters: `compute_theta()`


**Purpose:** Derive the association parameters $\theta$ 
 using event counts (death & progression) and total follow-up times.

**Main Inputs:**

-   `event_nums_trt`: Named numeric vector for the treatment arm in the form `c(ND=..., NP=..., Ns=...)` for OS, progression, and PFS endpoint counts, respectively.
-   `event_nums_ctr`: Similar vector for the control arm.
-   `fl_times`: Tibble of total follow-up times (from compute_followup()).
-   `ref`: (Optional) Reference group’s label in `fl_times$group`.

**Output:** A numeric vector of length 2:

-   `theta_trt`: Association parameter for treatment group.
-   `theta_ctr`: Association parameter for control group.

**Example Usage:**

```{r}
theta_vec <- compute_theta(
  event_nums_trt = c(ND = 135, NP = 144, Ns = 199),
  event_nums_ctr = c(ND = 174, NP = 193, Ns = 229),
  fl_times       = fl_times,
  ref            = "Placebo"
)
```

### 6. Compute win-loss probabilities: `compute_win_loss()`

**Purpose:** Calculate time-varying win/loss probabilities from OS and PFS, optionally using association parameters $\theta$.

**Main Inputs:**

-   `df_inc`: Data frame of increments (from compute_increments()).
-   `event_nums_trt`, event_nums_ctr: Named numeric vectors with death/progression counts.
-   `ref`: (Optional but important) Reference group label in `df_inc`. This determines how win-loss is defined.
-   `theta`: (Optional) Numeric vector from `compute_theta()`. If `NULL`, a simpler approach is used assuming independent endpoints, which may be inaccurate.


**Output:** A tibble with:

-   `time`: Time points.
-   `win_os`, `loss_os`: Win/loss from OS.
-   `win_pd`, `loss_pd`: Win/loss from PFS (when OS is tied).
-   `win`, `loss`: Overall win/loss.

**Example Usage:**

```{r}
df_wl <- compute_win_loss(
  df_inc         = df_increments,
  event_nums_trt = c(ND=135, NP=144, Ns=199),
  event_nums_ctr = c(ND=174, NP=193, Ns=229),
  ref            = "Placebo",
  theta          = theta_vec
)

```




## All-in-one approach: `run_win_loss_workflow()`

**Purpose:** Perform all the above steps in a single function call.

**Main Inputs:** (selected)

-   `os_file`, `pfs_file`: (Optional) CSV files for OS/PFS digitized data.
-   `os_data`, `pfs_data`: (Optional) Alternatively, already-loaded data frames for OS/PFS.
-   `event_nums_trt`, `event_nums_ctr`: Named numeric vectors of death/progression/PFS counts.
-   `ref_group`, `ref_group_labelled`: Reference group name before/after labeling.
-   `compute_theta`: If `TRUE`, also reads risk-table data and computes $\theta$.
-   `risk_table_file`, `risk_table_data`: (Optional) CSV or data frame with at-risk numbers.
-   `group_cols_at_risk`, `group_labels_at_risk`: Columns and labels for pivoting the risk table.

**Output:** A list containing:

-   `os_long`, `pfs_long`: Pivoted OS/PFS data.
-   `df_merged`: Merged OS/PFS tibble.
-   `df_increments`: The increments tibble.
-   `theta`: Computed $\theta$ parameters (if requested).
-   `df_wl`: Final win/loss tibble from `compute_win_loss()`.

**Example Usage:**

```{r}
# Example usage:
results <- run_win_loss_workflow(
  # Optionally specify input files or data frames
  os_file        = "camrelizumab_os.csv",
  pfs_file       = "camrelizumab_pfs.csv",

  # Event counts
  event_nums_trt = c(ND=135, NP=144, Ns=199),
  event_nums_ctr = c(ND=174, NP=193, Ns=229),

  # Reference group labeling
  ref_group        = "placebo",
  ref_group_labelled = "Placebo",

  # Merge OS/PFS & compute increments
  merge_os_pfs   = TRUE,

  # Compute theta using risk table
  compute_theta  = TRUE,
  risk_table_file   = "risk_table.csv",
  group_cols_at_risk   = c("placebo_chemo", "camr_chemo"),
  group_labels_at_risk = c("Placebo", "Camrelizumab")
)

# 'results' is a list:
# - results$os_long
# - results$pfs_long
# - results$df_merged
# - results$df_increments
# - results$theta
# - results$df_wl

```


## References
